---
title: PCA and beautiful biplot
author: "Daniel Hernández"
date: 2
categories: ["PCA"]
tags: ["PCA", "plot"]
---


<!--more-->

In this tutorial you will learn how to use the ``prcomp`` function to perform a principal component analyses (PCA) and display the results in a fully customizable format. I assume you are familiarized with the theory behind a PCA and have basic R-programming skills. 

## Example data (IRIS)

We are going to use the [iris](https://archive.ics.uci.edu/ml/datasets/iris) dataset as an example. It includes four numerical variables representing four flower characteristics and one categorical variable that corresponds to the taxonomic classification. 

```{r, include=T, echo=T, warning=F}
data(iris) # Load dataset
sapply(iris, class) #Check type of variables
```

## Objective

We want to know:


* Whether the variables that describe de flower can be summarized in two numeric axis or not.
* How variables are correlated among them.
* If different groups of plants have distinct flower characteristics


A good way to inspect the data is to analyse the correlation between variables. For that I like using the ``PerformanceAnalytics`` package. 

```{r}

PerformanceAnalytics:: chart.Correlation(iris[,-5],
                                         histogram=TRUE, 
                                         pch=10,
                                         method = "pearson")
```

There is a strong correlation between some variables, specially between ``Petal.Length`` and ``Petal.Width``. 


