---
title: PCA and beautiful biplot
type: book
highlight: true
author: "Daniel Hernández"
date: 2
categories: ["PCA"]
tags: ["PCA", "plot"]
---


<!--more-->

In this tutorial you will learn how to use the ``prcomp`` function to perform a principal component analyses (PCA) and display the results in a fully customizable format. I assume you are familiarized with the theory behind a PCA and have basic R-programming skills. 

We are going to use the [iris](https://archive.ics.uci.edu/ml/datasets/iris) dataset as an example. It includes four numerical variables representing four flower characteristics and one categorical variable that corresponds to the taxonomic classification. 

```{r, include=T, echo=T, warning=F}
data(iris) # Load dataset
sapply(iris, class) #Check the class of variables
```

## Preparing the data

A good way to inspect the data is to analyse the correlation between variables. For that I like using the ``PerformanceAnalytics`` package. 

```{r}

PerformanceAnalytics:: chart.Correlation(iris[,-5],
                                         histogram=TRUE, 
                                         pch=10,
                                         method = "pearson")
```

There are strong correlations between some variables, specially between ``Petal.Length`` and ``Petal.Width``. For this example we are going to keep both variables, but you could consider removing one when such strong correlations are detected. 

Next, we can visually explore the distribution of the variables.
```{r}
library(ggplot2)

iris.num <- iris[,-5] #variables to use in the PCA


ggplot(stack(data.frame(scale(iris.num))), aes(x=ind, y=values)) +
  geom_boxplot(width=0.5) +  #draw boxplot
  geom_jitter(width=0.1, alpha=0.5) #draw points 

```

The distribution seem symmetric enough. Otherwise, a proper transformation variables would be needed. 

## PCA

We can use the ``vegan`` package to perfoorm a PCA using the ``rda()``function.   

```{r, warning=FALSE, include=TRUE}

library(vegan)
pca.out <- rda(iris.num, scale=TRUE)
pca.out$inertia
```

The result of the PCA has been stored with the name of ``pca.out``. One important aspect of a PCA is the % of variance explained by each principal component, which can be consulted in the ``$CA$eig`` section of the PCA output. 

```{r}
kableExtra::kable(round(data.frame(summary(pca.out)[6]), 3))

varExp <- pca.out$CA$eig/sum(pca.out$CA$eig)*100 

```


```{r}
ggplot(data.frame(varExp,
                  PC= substr(names(varExp), 3,5)), 
       aes(x=as.numeric(PC), y=varExp))+
    geom_line(col="grey40", size=0.7)+
  geom_point(colour="coral3", size=3) + 
  theme(aspect.ratio=0.5)+
  xlab("PC")+
  ylab("Proportion explained (%)")

```


The first PC axis summarizes a big part of the variance (75%) and adding a second PC allows for includding more than 90% of the variance. 



