---
date: "2021-04-10"
highlight: true
title: Do biotic interactions affect ecosystem stability? 
subtitle: Simulating interaction networks in R. 
type: book
weight: 30
bibliography: mybibfile.bib
draft: false
hidden: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

htmltools::tagList(rmarkdown::html_dependency_font_awesome())

library(gridExtra)
library(ggplot2)
theme_set(theme_bw())
    theme_update(text=element_text(size=16, family= "serif"))
library(reshape2)
    library(igraph)
    library(plotly)
library(kableExtra)
library(doBy)
library(gganimate)
library(deSolve)
library(dplyr)
    
     theme_set(theme_bw())
    theme_update(text=element_text(size=12, family= "serif"),
                 panel.background = element_rect(fill = "white"), 
                 strip.background = element_rect(color="black", 
                                                 fill="transparent", size=1,
                                                 linetype="solid"),
                 legend.background = element_rect(fill = "transparent"),
                  plot.background = element_rect(fill = "transparent", color = NA),
     strip.text = element_text( color = "black", face = "bold.italic"),
     axis.ticks.length=unit(-0.15, "cm"), 
                 axis.text.x = element_text(margin=unit(c(0.5,0.5,0.5,0.5), "cm")), 
                 axis.text.y = element_text(margin=unit(c(0.5,0.5,0.5,0.5), "cm"))
     )

```


**When a natural ecosystem is free from perturbation it reaches a stable state where every species abundance stays the same. Or does it? **

The following video shows a simplified ecosystem simulation including individuals of 3 trophic levels (phytoplankton, zooplankton and fish). Each individual has an specified biomass that increases each time a prey is consumed, with part of the prey biomass being released as free nutrients. Stochastic death can also occur, in which case all the biomass is directly transformed into free nutrients. When a certain biomass is reached, individuals may reproduce themselves by splitting their biomass in two. Thus, the total mass of the "ecosystem" remains constant.


<center>
<iframe width="600" height="430" src="https://www.youtube.com/embed/G4qt5cSSf9Q" frameborder="0" allow="accelerometer; autoplay=1; clipboard-write; encrypted-media; gyroscope; picture-in-picture" allowfullscreen></iframe>
</center>

<br>

Even after hours of simulation there is nothing like an equilibrium being reached. Furthermore, periodic oscillations of predator-prey biomass can be observed. This behaviour resembles [Alfred J. Lotka](https://en.wikipedia.org/wiki/Alfred_J._Lotka) and [Vito Volterra](https://en.wikipedia.org/wiki/Vito_Volterra) model, which simulates two interacting species.

<style>
div.blue { background-color:#e6f0ff; border-radius: 5px; padding: 20px;}
</style>

<div class = "blue">

$$\frac{\text{d}P}{\text{d}t} = rP - cPZ$$

$$\frac{\text{d}Z}{\text{d}t} = ePZ - mZ$$

where:

* $P$ is the prey abundance
* $Z$ is the predator abundance
* $r$ is the intrinsic growth rate of the prey 
* $c$ is the effect of predation on the prey.
* $e$ is the effect of predation on the predator. 
* $m$ is the mortality rate of the predator


</div> 
<br> 
<br>

The model simulation shows a constantly oscillating system regardless of the initial conditions. In the plotted examples, the parameters are set to be $r$ = 0.2, $c$ = 0.3,  $e$ = 0.3,  $m$ = 0.3. 
```{r,incude=TRUE, echo=FALSE, fig.height = 2, out.width = 8, message= FALSE, warning=FALSE}

#Vector amb els paràmetres del model
par<-c(rp=0.2, 
       c=0.3, 
       e=0.3, 
       m=0.3)



#Interval de simulació (0-500) i resolució (1)
tint = 0.2
t<-seq(0, tint, tint)

P <- seq(0,3,0.2)
Z <- seq(0,3,0.2)

out.list <- list()

n = 0

#funció del model
model_1<-function(t, init,par){
   with(as.list(c(init,par)),{
     
       #Sistema d'equacions diferenciasl 
      dP<- rp*P - c*P*Z 
      dZ<- e*P*Z - m*Z
      
      list(c(dP,dZ))
      })
  
}

for(i in P){
  for(j in Z){
    n = n + 1
    #Vector amb els valors inicials de les variables d'estat
    init<-c(P=i,
            Z=j)
    
        
    out<-as.data.frame(ode(init,t, model_1,par))
    out<- out[(out$time== tint),]
    
    out$Pin <- i
    out$Zin <- j
    
    out.list[[n]] <- out

  }
}


out.all <-bind_rows(out.list)
out.all$size <- sqrt(abs(out.all$P - out.all$Pin)^2 +abs(out.all$Z - out.all$Zin)^2) 

out.all$vP <- (out.all$P - out.all$Pin)/out.all$size
out.all$vZ <- (out.all$Z - out.all$Zin)/out.all$size 

```


```{r,incude=TRUE, echo=FALSE, message= FALSE, warning=FALSE}

#Interval de simulació (0-500) i resolució (1)

t<-seq(0, 35, 0.1)

P <- seq(1,3,0.5)
Z <- 0.7

outTime.list <- list()

n = 0

for(i in P){
  for(j in Z){
    n = n + 1
    #Vector amb els valors inicials de les variables d'estat
    init<-c(P=i,
            Z=j)
        
    out<-as.data.frame(ode(init,t, model_1,par))
    
    out$Pin <- i
    out$Zin <- j
    out$run <- n
    
    outTime.list[[n]] <- out

  }
}


outTime.all <-bind_rows(outTime.list)
  
ggplot(outTime.all, aes(x=P, y=Z)) +
  theme(aspect.ratio = 1,
        panel.grid = element_blank(),
        panel.background= element_blank(),
        plot.background = element_blank(),
        legend.position = "none")+
  geom_segment(data=out.all, aes(x = Pin, y = Zin, xend = Pin+vP*0.1, yend = Zin + vZ*0.1, col=size),
                  arrow = arrow(length = unit(0.2, "cm")),size=1)+
  scale_color_gradientn(colours=c("green","yellow", "orange", "coral2", "purple"))+
  scale_size_continuous(range = c(0.5,1.5))+ xlim(c(0,3.1)) + ylim(c(0,2.8))+
  geom_path(aes(group=run), size=1) + 
  ggrepel::geom_label_repel(data=data.frame(P=seq(1,3,0.5), 
                                            Z=rep(0.7,5), 
                                            label=paste(seq(1,3,0.5),",", rep(0.7,5))),
                            aes(x=P, y=Z, label=label),
                            force_pull = 0.9, force=10)




```



## Simulating an interacting community


To study the effect of biotic interactions when a larger number of organisms is involved, a simplified interacting community can be simulated as follow: 


<div class = "blue">


$$\frac{\text{d}C_{i}}{\text{d}t} = r_iC_{i}\left( 1- \frac{C_{i} - [\alpha_{ji}] C_{j}}{k_i}\right)$$
Where:

* $C_{i}$ is a vector with the abundance of each species.
* $t$ represents time.
* $r_{i}$ is a vector that represents the intrinsic growth rate of each species.
* $\alpha_{ij}$ is a matrix that contains every pairwise interaction.
* $K_i$ is a vector that represents the carrying capacity for each species.

</div>

<br> 

An example of interaction matrix $\alpha_{ij}$ looks like this:
  
```{r, echo=FALSE}

 Int<- read.csv("network.csv")[,-1]

row.names(Int)<- names(Int)
 
knitr::kable(Int[1:10,1:10]) %>%   kable_styling(font_size = 12)
  
``` 

In this example there are predator-prey, mutualistic and competitive interactions among 10 species (Vn). For instance, V1 has:

* A trophic interaction (+, -) of weight 0.2 with V2
* A competence interaction (-, -) of weight 0.05 with V3
* A mutualistic interaction (+,`+) of asymetric weight (0.20, 0.05) with V9. 

Now we can see how different aspects of biotic interactions affect the system stability. 
<br>

# Connectance 

An interacting community can be represented as a network. The nodes of the network represent organisms and the edges represent interactions. The _connectance_ is a measure of how densely interconnected the network is. Consequently, a higher number of interactions for a given number of nodes increases the network connectance. 


```{r, echo=FALSE, fig.height= 5, dpi=300}

op <- par(mfrow = c(1, 2)) 
par(mar=c(0, 0, 0, 0))
Int_lc<- read.csv("network_lc.csv")[,-1]
network <- simplify(graph.adjacency(abs(as.matrix(Int_lc)), mode= "lower", weighted= TRUE))

plot(
     network, 
     vertex.label=NA,
     vertex.color="lightblue",
     layout=layout.circle,
     vertex.size=20, #Edge size related to the degree in the network
     edge.width = E(network)$weight*15, #Edge width related to the weight
     edge.curved=F) #F
title("Low C", line=-10)



Int_hc<- read.csv("network_hc.csv")[,-1]
network <- simplify(graph.adjacency(abs(as.matrix(Int_hc)), mode= "lower", weighted= TRUE))

plot(
     network, 
     vertex.label=NA,
     vertex.color="lightblue",
     layout=layout.circle,
     vertex.size=20, #Edge size related to the degree in the network
     edge.width = E(network)$weight*15, #Edge width related to the weight
     edge.curved=F) #F
title("High C", line=-10)


op <- par(mfrow = c(1, 1)) 
```


Random interaction matrices have been generated in the connectance gradient and the equilibriums have been calculated. The eigenvalues obtained from the diagonalization of the Jacobian matrix has been studied to asses the probability of stable equilibrium at increasing connectance scenarios. 

```{r, message=F, include=T, echo=F, warning=FALSE, dpi=200, fig.width=9, fig.height=5}

stability.data <- read.csv("nInt.data.csv")


ggplot(stability.data, aes(y=as.numeric(stability), x=int)) + geom_jitter(width=0.5, height = 0.02,size=3, alpha=0.1)+ 
    geom_smooth(method="glm",method.args=list(family=binomial), colour="coral", se=F )+ ylab("Stability")+ xlab("Connectance (Interactions/Species)")+theme(legend.position="") 
```

The results show a decreasing probability of stability in the connectance gradient, which may reflect a destabilizing nature of biotic interactions. 


# Interaction Strength

To keep it simple, the strength of biotic interactions have been restricted to two possible values: _weak_ (0.05 or -0.05) and _strong_ (0.20 or -0.20). In the previous simulations, the proportion of weak and strong interactions was constant. However, we will now asses the effect of increasing the proportion of strong interactions over the probability of stable equilibrium. 


```{r, echo=FALSE, fig.height= 5, dpi=300}

op <- par(mfrow = c(1, 2)) 
par(mar=c(0, 0, 0, 0))
Int_wI<- read.csv("network_wI.csv")[,-1]
network <- simplify(graph.adjacency(abs(as.matrix(Int_wI)), mode= "lower", weighted= TRUE))

plot(
     network, 
     vertex.label=NA,
     vertex.color="lightblue",
     layout=layout.circle,
     vertex.size=20, #Edge size related to the degree in the network
     edge.width = E(network)$weight*15, #Edge width related to the weight
     edge.curved=F) #F
title("Low S/W", line=-10)



Int_sI<- read.csv("network_sI.csv")[,-1]
network <- simplify(graph.adjacency(abs(as.matrix(Int_sI)), mode= "lower", weighted= TRUE))

plot(
     network, 
     vertex.label=NA,
     vertex.color="lightblue",
     layout=layout.circle,
     vertex.size=20, #Edge size related to the degree in the network
     edge.width = E(network)$weight*15, #Edge width related to the weight
     edge.curved=F) #F
title("High S/W", line=-10)


op <- par(mfrow = c(1, 1)) 
```



```{r, message=F, include=T, echo=F, warning=FALSE, dpi=200, fig.width=9, fig.height=5}

stability.data <- read.csv("propSInt.data.csv")




 ggplot(stability.data, aes(y=as.numeric(stability), x=sIntProp)) + geom_jitter(width=0.025, height = 0.02,size=3, alpha=0.1)+ 
     geom_smooth(method="glm",method.args=list(family=binomial), colour="coral", se=F )+ ylab("Stability")+ xlab("Strong Interactions / weak interactions")+theme(legend.position="")


```

It seems like increasing the proportion of strong interactions also has a destabilizing impact. 
<br>



When calculating the probability of stable equilibrium for different combinations of interaction strengths and network connectances we obtain the following result: 

```{r, message=F, echo=F, include=T, dpi=300, fig.width=4.5, fig.height=2}

library(ggtern)
library(doBy)


stability.data <- read.csv("nInt_sIntProp.data.csv")
stability.data.sum <- summaryBy(stability+ oscilations+ resilience ~ nInt + sIntProp , FUN= c(mean), data=stability.data)

library(gridExtra)

grid.arrange(
  ggplot(stability.data.sum, aes(x=nInt, y=sIntProp, z= stability.mean))+
               theme_bw()+
  stat_summary_hex(bins=10,col="white", size=0.5)+ 
    scale_fill_gradientn(name= "Stability",
                         colors= c("coral4","wheat1","deepskyblue4"))+ 
    xlab("Interactions/Species")+ylab("sI/wI")+
        theme_classic()+
    theme(text=element_text(size=12, family= "serif"), aspect.ratio = 1, axis.line = element_blank()), ncol=1)




```




## Interaction Type

Finally, we will asses how the type of interaction affects the system stability. To do so, random interaction matrices of consistent properties except for the proportion of each interaction type will be generated. I have covered as much combinations as possible while keeping my laptop from burning. 







